Epistemic Semantics in Guarded String Models

### 2020-09-18 tex/draft.tex
tex edits

\section{Introduction and Related Work}
Linguistic semantics in the Montague tradition proceeds by assigning
propositional {\em semantic values} to disambiguated sentences of a
natural language.  A proposition is a set or class of {\em possible
  worlds}.  These are often assumed to be things with the same
nature and complexity as the world we occupy \citep{lewis1986plurality}.  But
alternatively, one can work with small idealized models, in order to
illustrate and test ideas.  To build such models, spaces of worlds and
individuals are stipulated as small finite sets, and semantic values
of lexical items are constructed as functions or relations from these
small sets.  Such toy or idealized models are useful in research and
in teaching, in that it is possible to represent propositions finitely
and explicitly, and to calculate with them. The point of this paper is
to scale up toy or idealized models to countable sets of worlds, and
to constructive and computable modeling of epistemic alternatives for
agents.  We describe a certain systematic way of defining such models,
and illustrate how to apply them in natural language semantics.
The focus is on epistemic semantics and clausal embedding. The fundamental move is to
identify possible worlds with strings of primitive events, so that
propositions are sets of strings. An advantage in this is that it
allows for a mathematical description of an algebra of propositions,
coupled with a computational representation using either lazy lists of
strings, or finite state machines that describe sets of strings.

The approach taken here synthesizes five antecedents in a certain way.
John McCarthy's {\em Situation Calculus} is the source of the idea of
constructing possible worlds as event sequences
\citep{mccarthy1963situations,reiter2001knowledge}.  The algebraic
theory of {\em Kleene Algebra with Tests} characterizes algebras with
elements corresponding to propositions and event types in our
application \citep{kozen2001automata}.\marginpar{2001 or 2003?}  The
models we propose are an epistemic extension of guarded string models
for KAT, where a unary operation interpreted as an existential
epistemic modality is included for each agent.  {\em Action models} in
dynamic epistemic semantics introduce the technique of constructing
epistemic models from primitive alternative relations on events, in
order to capture the epistemic consequences of perceptual and
communicative events \citep{baltag1999logic}. This is the basis for
our construction of epistemic alternative relations.  Literature on
finite state methods in linguistic semantics has used event strings
and sets of event strings to theorize about tense and aspect in
natural language semantics
\citep{fernando2004finite,fernando2007observing,carlson2009tense} and
to express intensionality \cite{fernando2017intensions}.  Work on
finite state intensiononal semantics investigates how to do the
semantics of intensional complementation, including indirect
questions, in a setting where compositional semantics is expressed in
a finite state calculus \citep{rooth2017finite,collard2018finite}.  We
adopt this in our syntax-semantics interface for English.

@@

\begin{example}   \label{gs1}
  \renewcommand{\arraystretch}{1.2}
  \begin{tabular}[t]{clc}
{\textit{World}} & {\textit{Guarded string}} & \textit{Length}\\ \hline
$v_1$ & $\hat 0 \,u\, \hat 1 \,u\, \hat 2 \,d\, \hat 1 \,u\, \hat 2$ & $4$ \\
$v_2$ & $\hat 0 \,u\, \hat 1 \,u\, \hat 2 \,u\, \hat 3$ & $3$ \\   
$w_1$ & $H$ & $0$\\
$w_2$ & $H \,a_0\, H$ & $1$ \\  
$w_3$ & $T\,a_0\, T \,b_0\, T$ & $2$\\
  \end{tabular}
    \renewcommand{\arraystretch}{1}
\end{example}

@@

\section{Translation into the finite state calculus}
The finite state calculus is an algebra of regular sets of strings and regular
relations between strings that was designed for use in computationonal phonology and
morphographemics \citep{kaplan1994regular,beesley2003finite}.  Current
implementations allow for the definition of functions that have the status of defined
operations on regular sets and relations \citep{hulden2009foma,karttunen2010update}.
Such definitions are used here to construct of a model for epistemic KAT inside the
finite state calculus.  The space of worlds is a set of ordinary strings.  Bit
sequences (sequences of 0's and 1's) are used for atoms, and these alternate with
event symbols in the encoding of a world. \ex{gs1fst} gives the encoding of some of
the worlds from the examples.  A string is a finite sequence of symbols, and {\tt 0},
{\tt 1}, {\tt u}, and {\tt d}, are symbols.  {\tt a0} and {\tt b0} are special
multi-character symbols that are found in implementations of the finite state
calculus.

@@

 \hspace{-13mm} \begin{tabular}[t]{cl}
{\textit{World}} & {\textit{String}} \\
% $v_1$ & $ 0 \,u\,  1 \,u\,  2 \,d\,  1 \,u\,  2$ \\
$v_1$ & {\tt 0 0 u 0 1 u 1 0 d  0 1 u  1 0} \\
$v_2$ & $ 0 \,u\,  1 \,u\,  2 \,u\,  3$ \\   
$w_1$ & $1 0$ \\
$w_2$ & $1 0 \,a_0\, 1 0$ \\  
$w_3$ & $0 1 \,a_0\, 0 1 \,b_0\, 0 1$ \\
  \end{tabular}
    \renewcommand{\arraystretch}{1}
\end{example}

Operations such as KAT product are defined operations in the finite
state calculus.

of epistemic KAT in a
string algebra. The methodology follows Section 2 closely.  Let ${\cal K}$ be an
epistemic algebra as described in Section 2.  A given element of ${\cal K}$ is
represented in the the string algebra by the very same set of strings, i.e. by a set
of strings that have the form of a sequence of bare event symbols, with interleaved
Boolean vectors. Product in the KAT can not be modeled as concatenation in the string
algebra, because this would not enforce identity of states, and would result in
lengthening Boolean vectors at the concatenation point.  Instead, KAT product and KAT
Kleene star are defined operations in the string algebra, see Figure \ref{fstfigure1}.  The
operations concatenate in the string algebra, delete strings with non-matching tests,
and then delete the second of two tests create a well-formed guarded string.

@@

\begin{hangingpar}
\begin{verbatim}
define Cn(X,Y) 
 [[[X Y] & Wf0] .o. Squash].l;
\end{verbatim}
KAT product in Fst, where {\tt \&} is intersection, {\tt .o.} is relation composition,
and {\tt .l} is relation image.
\end{hangingpar}

@@

\begin{hangingpar}
\begin{verbatim}
define Kst(X) St |  Kpl(X);
\end{verbatim}
KAT Kleene plus and Kleene star in Fst. The Fst operation {\tt |} is union.
\end{hangingpar}

\caption{Translation into Fst of KAT product and KAT Kleene star.}
\label{fstfigure1}
\end{figure}

@@

\begin{tabular}{ll}
  $a$ & Relational Kleene plus in the string algebra \\
  $b$ & Constrain domain and co-domain to contain \\
      & no unmatched tests. \\
  $c$ & Reduce doubled tests to a single \\
      & test in the domain and co-domain. \\
      & {\tt{Squash.i}} is the inverse \\
      & of {\tt{Squash}}.
\end{tabular}

\begin{hangingpar}
{\tt{define Kst(R) [St\hspace{1pt}.x.\hspace{1pt}St]\hspace{1pt}|\hspace{1pt}Kpl(X);}}
The Fst operation {\tt{.x.}} is Cartesian product.
\end{hangingpar}

@@

An architecture of interpretation by transtlation is employed, where English
sentences are mapped to terms in the logical language (--) via an interpreted
grammar, and these terms are in turn interpreted as propositions (sets of possible
worlds). For the latter, there are options of translation into the finite state
calculus in order to represent propositions as finite state machines (Section 3), and
representation in Haskell via lazy lists of guarded strings (Section 4).  The grammar
is a semantically interpreted multi-modal categorial grammar, consisting of a lexicon
of words, their categorial types, and interpretations in a logical lambda language.
Figure \ref{phenomenafigure} lists phenomena that are covered.

@@

As illustrated towards the end, there is recursion through conjuction and verbal complementation,
so that the language is infinite, and includes talk of beliefs about beliefs, or in general, talk
of arbitrarily iterated belief.

Figure \ref{lexiconfigure} gives illustrative lexical entries. The grammar and semantics are in
certain way optimized for a simple fragment of English concerned with clausal
complementation.  The agent names {\em Amy} and {\em Bob} contribute the epistemic
alternative relations for those agents, rather than individuals.  This is possible
because the agents are never arguments of extensional predicates, so what matters
about the agents is their epistemic alternative relations.  The root verb {\em know}
contributes existential modal force.  The complementizers {\em that} and {\em
  whether} are the heads of their dominating clauses, and assemble an alternative
relation, modal force, and proposition contributed by the complement.  These
complementizers introduce the dual via two negations, in order to arrive universal
modal force.  These moves are offered here as a way of constructing a compact
interpreted grammar. They can easily be reformulated in a more comprehensive
interpreted grammar of English.

\begin{figure}  
\begin{tabular}[t]{l|c|c}
Amy & $e$ & ${R}_a$\\
Bob & $e$ & ${R}_b$\\
it & $d$ & $d$ \\
heads & $d \backslash_D t$ & $\lambda x.0^c{\cdot}h$ \\
tails & $d \backslash_D t$ & $\lambda x.0^c{\cdot}!h$ \\
is & $(d \backslash t)\slash(d \backslash_D t)$ & $\lambda P.\lambda x.P x$ \\
knows & $(e \backslash t)\slash_M t$ &  $\lambda p.\lambda R. {\Diamond}R p$\\
that  & \multicolumn{2}{l}{$(((e \backslash t)\slash_M t) \backslash (e \backslash t)) \slash t$} \\
& \multicolumn{2}{l}{$\lambda p.\lambda m.\lambda R.{\sim}(m ({\sim}p) R)$} \\
whether  & \multicolumn{2}{l}{$(((e \backslash t)\slash_M t) \backslash (e \backslash t)) \slash t$} \\
& \multicolumn{2}{l}{$\lambda p.\lambda m.\lambda R.{\sim}(m ({\sim}p) R)$} \\
& \multicolumn{2}{r}{$+ {\sim}(m p R)$} \\
\end{tabular}
\caption{Partial categorial grammar lexicon. The first column has a word form.
  the second column a categorial type, and third column a semantic translation in
a logical language that extends the Epik term language with lambda.}
\label{lexiconfigure}
\end{figure}

Multimodal categories such as $\backslash_D$ and $\backslash_M$ are used to control
the derivation.  For instance the category of {\em heads} $d \backslash_D t$.  The
dummy expletive subject {\em it} has category $d$, but the phrase {\em it heads} of
category $t$ can not be formed, because $\backslash_D$ is not syntactically active as
a function.  Instead {\em it is heads} can be formed with a predicator {\em is} of
category $(d \backslash t)\slash(d \backslash_D t)$.  (This uses Lambek/Bar-Hillel
notation for slashes, so that $(d \backslash t)\slash(d \backslash_D t)$ combines
with $d \backslash_D t$ on the right to give a value that combines with $d$ on the
left to give $t$.)  Similarly {\em knows} has a category with the top-level slash
$\slash_M$, and combines to form a sentence as an argument of {\em that} or {\em
  whether}, which has a category that looks for the category of {\em know} on the
left, after combining with a complement sentence on the right.

The semantic translations in the third column of Figure \ref{lexiconfigure} use the Epik term
language, incremented with lambda.  The body of $\lambda x.0^c{\cdot}h$, which is the
semantic lexical entry for {\em heads}, is a term denoting the set of all worlds
where the coin is heads, represented as the set of all guarded strings that end with
a Boolean valuation where the primitive proposition $h$ (it's heads) is true.  There
is $\lambda x$ at the front because of a correspondence the grammar formalism uses a
correspondence between syntactic and semantic types. However, it does not bind
anything, because sentences such as {\em it isn't heads} have an expletive subject.
The body ${\Diamond}R p$ of $\lambda p.\lambda R. {\Diamond}R p$, which is the
semantic lexical entry of {\em knows}, is an Epik term denoting the pre-image of the
set of worlds $p$ according to the relation $\doublehat{R}$ between guarded strings
that is determined by the event-level relation $R$.  This is not the right semantics
for {\em Amy knows that it's heads}, because it is an existential modality ${\Diamond_R}p$,
rather than an universal modality ${\Box_R}p$.  This is corrected by the complementizer
{\em that} or {\em whether}, which introduces the dual.

Sentences are parsed with a chart parser for categorial grammar.  The semantics for
complex phrases are obtained by syntactic application of semantic translations,
accompanied by beta reduction.  Semantic terms in the parsing formalism are
expressions of untyped lambda calculus.  The grammar is set up so that lambda is
eliminated by beta reduction in the semantic term corresponding to a sentence. In
consequence, the semantic term translating a sentence is a term of the Epik language
(--).  Such a term designates a set of possible words (guarded strings) in the
possible worlds model deterimened by an Epik specification such as the one in Figure
1.  \exi{three}{a} is an English sentence with predicate conjunction and three levels
of clausal embedding.  Using the grammar and parser, the sentence is mapped to the
Epik term \exi{b}.  Using the result from Section 3, this term can be mapped in an
implementation of the finite state calculus to a finite state machine that represents
a countably infinite set of possible worlds, represented as guarded strings.  Using
the result from Section 4, it can be mapped to an infinite lazy list of guarded
strings, representing the same set of possible worlds.  Either of these is a concrete
computational representation of the propositional semantic value \semval{Amy knows
  that Bob knows that Amy knows whether it is heads and knows that Bob does not know
  that Amy knows that it is tails}, in the familiar sense of Montague semantics for
natural language.

\begin{examples}  \label{three}
\itema  Amy knows that Bob knows that Amy knows whether it is heads and
knows that Bob does not know that Amy knows that it is tails.
\itemb
\end{examples}

@@

>>>>

\section{Introduction and Related Work}
Linguistic semantics in the Montague tradition proceeds by assigning
propositional {\em semantic values} to disambiguated sentences of a
natural language.  A proposition is a set or class of {\em possible
  worlds}.  These are often assumed to be things with the same nature
and complexity as the world we occupy \citep{lewis1986plurality}.  But
alternatively, one can work with small idealized models, in order to
illustrate and test ideas.  To build such models, spaces of worlds and
individuals are stipulated as small finite sets, and semantic values
of lexical items are constructed as functions or relations from these
sets.  Such toy or idealized models are useful in research and in
teaching, in that it is possible to represent propositions finitely
and explicitly, and to calculate with them. The point of this paper is
to scale up toy or idealized models to countable sets of worlds, and
to constructive and computable modeling of epistemic alternatives for
agents.  We describe a certain systematic way of defining such models,
and illustrate how to apply them in natural language semantics.  The
focus is on epistemic semantics and clausal embedding. The fundamental
move is to identify possible worlds with strings of primitive events,
so that propositions are sets of strings. An advantage in this is that
it allows for a mathematical description of an algebra of
propositions, coupled with a computational representation using either
lazy lists of strings, or finite state machines that describe sets of
strings.

The approach taken here synthesizes five antecedents in a certain way.
John McCarthy's {\em Situation Calculus} is the source of the idea of
constructing possible worlds as event sequences
\citep{mccarthy1963situations,reiter2001knowledge}.  The algebraic
theory of {\em Kleene Algebra with Tests} characterizes algebras with
elements corresponding to propositions and event types in our
application \citep{kozen2001automata}. The models we propose are an
epistemic extension of guarded string models for KAT, where a unary
operation interpreted as an existential epistemic modality is included
for each agent.  {\em Action models} in dynamic epistemic semantics
introduce the technique of constructing epistemic models from
primitive alternative relations on events, in order to capture the
epistemic consequences of perceptual and communicative events
\citep{baltag1999logic}. This is the basis for our construction of
epistemic alternative relations.  Literature on {\em finite state
  methods in linguistic semantics} has used event strings and sets of
event strings to theorize about tense and aspect in natural language
semantics
\citep{fernando2004finite,fernando2007observing,carlson2009tense} and
to express intensions \cite{fernando2017intensions}.  Work on {\em
  finite state intensiononal semantics} has investigated how to do the
semantics of intensional complementation, including indirect
questions, in a setting where compositional semantics is expressed in
a finite state calculus \citep{rooth2017finite,collard2018finite}.  We
adopt this in our syntax-semantics interface for English.

@@

\begin{example}   \label{gs1}
  \renewcommand{\arraystretch}{1.2}
  \begin{tabular}[t]{clc}
{\textit{World}} & {\textit{Guarded string}} & \textit{Length}\\ \hline
$v_1$ & $\hat 0 \,u\, \hat 1 \,u\, \hat 2 \,d\, \hat 1 \,u\, \hat 2$ & $4$ \\
$v_2$ & $\hat 0 \,u\, \hat 1 \,u\, \hat 2 \,u\, \hat 3$ & $3$ \\   
$w_1$ & $H$ & $0$\\
$w_2$ & $H \,a_1\, H$ & $1$ \\  
$w_3$ & $H\,a_1\, H \,b_1\, H$ & $2$\\
  \end{tabular}
    \renewcommand{\arraystretch}{1}
\end{example}

@@

\section{Translation into the finite state calculus}
The finite state calculus is an algebra of regular sets of strings and regular
relations between strings that was designed for use in computationonal phonology and
morphographemics \citep{kaplan1994regular,beesley2003finite}.  Current
implementations allow for the definition of functions that have the status of defined
operations on regular sets and relations \citep{hulden2009foma,karttunen2010update}.
Such definitions are used here to construct of a model for epistemic KAT inside the
finite state calculus.  The space of worlds is a set of ordinary strings.  Bit
sequences (sequences of 0's and 1's) are used for atoms, and these alternate with
event symbols in the encoding of a world. \ex{gs1fst} gives the encoding of some of
the worlds from the examples.  A string is a finite sequence of symbols, and {\tt 0},
{\tt 1}, {\tt u}, and {\tt d}, are symbols.  {\tt a0} and {\tt b0} are
multi-character symbols that are found in implementations of the finite state
calculus. The multi-character symbol {\tt a0} is an element of the alphabet that has
no connection with the element {\tt a}.

@@

 \hspace{-13mm} \begin{tabular}[t]{cl}
{\textit{World}} & {\textit{String}} \\
% $v_1$ & $ 0 \,u\,  1 \,u\,  2 \,d\,  1 \,u\,  2$ \\
$v_1$ & {\tt 0 0 u 0 1 u 1 0 d  0 1 u 1 0} \\
$v_2$ & {\tt 0 0 u 0 1 u 1 0 u 1 1} \\
$w_1$ & {\tt 1 0} \\
$w_2$ & {\tt 1 0 a1 1 0} \\
$w_3$ & {\tt 1 0 a1 1 0 b1 10} \\  
  \end{tabular}
    \renewcommand{\arraystretch}{1}
\end{example}

Terms in the finite state calculus are interpreted as sets of strings, or for
relational terms, as relations between strings.  Computationally, the sets and
relations are represented by finite state acceptors.  As used here, a program in the
Fst language of the finite state calculus is a straight-line program that defines a
sequence of constants naming sets, constants naming relations, and functions mapping
one or more regular relations to a regular set or relation.  The functions are
defined as macros.  To illustrate, \ex{stex1} is part of an Epik program that
declares states and a state constraint.  This is equivalent to the Fst program
\ex{stex2}. The result of the sequence of definitions is that {\tt St} is the set of
Boolean vectors or length four that are well-formed according to the state formula;
Nst is the complement operation for tests; H1 is the set of well-formed Boolean
vectors that have 1 in the first position; and so forth.  Translation from an Epik
program to an Fst program is accomplished in Haskell.

\begin{example}\label{stex1}
Part of an Epik program
\end{example}

\vspace{-3mm} {\small
\begin{verbatim}  
   state H1 T1 H2 T2
   constraint !(H1 T1) & (H1 + T1) 
              !(H2 T2) & (H2 + T2) 
\end{verbatim}}

\begin{example}\label{stex2}
Corresponding Fst program
\end{example}

\vspace{-3mm}
{\small 
\begin{verbatim} 
   define Bool ["0"|"1"];
   define St0 [Bool Bool Bool Bool];
   define H1 [1 Bool Bool Bool];
   define T1 [Bool 1 Bool Bool];
   define H2 [Bool Bool 1 Bool];
   define T2 [Bool Bool Bool 1];
   define St St0 & ((((St0 - (H1 & T1)) 
                 & (H1 | T1)) 
                 & (St0 - (H2 & T2))) 
                 & (H2 | T2));
   define Nst(X) [St - X];
   define H1 St & H1;
   define T1 St & T1;
   define H2 St & H2;
   define T2 St & T2;
\end{verbatim}}

Event symbols are interpreted as event types, in the way discussed in connection with
\ref{events1}.  For instance, defines an event type {\tt paT2type} of Amy peeking at the second coin
in a scenario with two coins, and seeing that it is tails.\footnote{After the definition,
  {\tt paT2type} is a constant symbol denoting a set of two strings. On the right hand side of the definition,
  {\tt paT2} is the unit set of the symbol {\tt paT2}.  
}  The first conjunct
expresses that this event can only happen if the second coin is tails ({\tt H2}), and that the
state of the second coin does not change.  The second conjunct expresses that the
event does not change the state of the first coin -- either it is heads ({\tt H1}) and
remains heads, or it is tails ({\tt T1}) and remains tails.  Such formulas are mapped from
effect formulas in Epik programs. The state formula is imposed as well, since it
figures in the definition of state constants such as {\tt H1}.

\begin{example} \label{peek1}
{\small 
\begin{verbatim}
define paT2type [[[T2 paT2] T2] & 
  [[[H1 paT2] H1] | [[T1 paT2] T1]]];
\end{verbatim} }
\end{example}

Semantically in Fst, {\tt paT2type} is a set of two strings, which can be displayed
in an interpreter for the language by reading {\tt paT2type} as a regular expression
and printing its extension, see \ex{printwords1}.

\begin{example}
\begin{verbatim}
xfst[1]: regex paT2type;
3.0 Kb. 16 states, 16 arcs, 2 paths.
xfst[2]: print words
0 1 0 1 paT2 0 1 0 1
1 0 0 1 paT2 1 0 0 1
\end{verbatim}
\end{example}

The finite state calculus has a product operation of string concatenation raised to sets.
Concatenation of strings with atoms (Boolean vectors) at both ends has the effect of
doubling atoms at the juncture, and does not enforce matching of atoms at the juncture.  Therefore
KAT product can not be identified with product in the finite state calculus. Instead,
KAT product and KAT Kleene star are defined operations, see Figure \ref{fstfigure1}.
The binary product operation {\tt Cn} and the unary Kleene star operation
{\tt Kst} combine strings in the string algebra, remove strings with
non-matching atoms, and then delete the second of two tests to create well-formed
guarded strings.  Matching of atoms is enforced with the set {\tt Wf0}, which is
the set of strings that do not contain non-matching Boolean vectors.  The containment
operator (expressed by the dollar sign) and complement (expressed by tilde) are operators of the finite
state calculus.  The set of non-matching sequences of atoms {\tt UnequalStPair} is defined by a finite disjunction.
Deletion is accomplished by a re-write rule in the finite state calculus, which is a
notation for defining regular relations by contextually constrained substitutions
(REFS).  In this case, is {\tt Squash} us a regular relation that deletes an atom
(a sequence of 0's and 1's of a certain length) when it follows an atom.\footnote{
  This is a non-equal length regular relation. The finite state calculus includes such
  relations, and they can be used with relation composition and relation domain and co-domain.
  They are restricted in that the complement and set difference for non-equal length relations is not defined.
The epistemic alterative relations that are defined in Figure -- are equal-length relations.
  }

@@

\begin{hangingpar}
\begin{verbatim}
define Cn(X,Y) 
 [[[X Y] & Wf0] .o. Squash].l;
\end{verbatim}
KAT product.
\end{hangingpar}

@@

\begin{hangingpar}
\begin{verbatim}
define Kst(X) St |  Kpl(X);
\end{verbatim}
KAT Kleene plus and Kleene star. The Fst operation {\tt |} is union.
\end{hangingpar}

\caption{\raggedright Definition in Fst of KAT product and KAT Kleene star.
  Where {\tt X} and {\tt Y} are regular sets and
  {\tt R} and {\tt S} are regular relatiions,
  {\tt X{\&}Y} is the intersection of {\tt X} and {\tt Y},
  {\tt X|Y} is the union of {\tt X} and {\tt Y},
  {\tt \~{}X} is the complement of {\tt X},
  {\tt R{.o.}S} is the composition of {\tt R} and {\tt S},
  {\tt R{.}l} is the co-domain of {\tt R}, and
  {\tt {\$}X} is the set of strings that have a substring in $X$.   
}
\label{fstfigure1}
\end{figure}

@@

\begin{tabular}{ll}
  $a$ & Relational Kleene plus in the string algebra \\
  $b$ & Constrain domain and co-domain to contain \\
      & no unmatched tests. \\
  $c$ & Reduce doubled tests to a single \\
      & test in the domain and co-domain. \\
\end{tabular}

\begin{hangingpar}
{\tt{define Kst(R)
    [St\hspace{1pt}.x.\hspace{1pt}St]\hspace{1pt}|\hspace{1pt}Kpl(X);}} The Fst
operation {\tt{.x.}} is Cartesian product.  {\tt{R.i}} is the inverse of relation
{\tt{R}}.
\end{hangingpar}

@@

An architecture of interpretation by transtlation is employed, where
English sentences are mapped to terms in the logical language via an
interpreted grammar, and these terms are in turn interpreted as
propositions (sets of possible worlds). For the latter, there are
options of translation into the finite state calculus in order to
represent propositions as finite state machines (Section 4), and
representation in Haskell via lazy lists of guarded strings (Section
5).  The grammar is a semantically interpreted multi-modal categorial
grammar, consisting of a lexicon of words, their categorial types, and
interpretations in a logical lambda language.  Figure
\ref{phenomenafigure} lists phenomena that are covered.  There is
recursion through conjuction and verbal complementation, so that the
language is infinite. It includes talk of beliefs about beliefs, or in
general, talk of arbitrarily iterated belief.

@@

Figure \ref{lexiconfigure} gives illustrative lexical
entries.\footnote{Category symbols use Lambek/Bar-Hillel notation for
  slashes, so that $(d \backslash t)\slash(d \backslash_D t)$ combines
  with $d \backslash_D t$ on the right to give a value that combines
  with $d$ on the left to give $t$.} The grammar and semantics are
optimized for a simple fragment of English concerned with
clausal complementation.  The agent names {\em Amy} and {\em Bob}
contribute the epistemic alternative relations for those agents,
rather than individuals.  This is possible because the agents are
never arguments of extensional predicates, so what matters about the
agents is their epistemic alternative relations.  The root verb {\em
  know} contributes existential modal force.  The complementizers {\em
  that} and {\em whether} are the heads of their dominating clauses,
and assemble an alternative relation, modal force, and proposition
contributed by the complement.  These complementizers introduce the
dual via two negations, in order to express universal modal force.
These moves are offered here as a way of constructing a compact
interpreted grammar. They can easily be reformulated in a more
comprehensive interpreted grammar of English.

\begin{figure}  
\begin{tabular}[t]{l|c|c}
Amy & $e$ & ${R}_a$\\
Bob & $e$ & ${R}_b$\\
it & $d$ & $d$ \\
heads & $d \backslash_D t$ & $\lambda x.0^c{\cdot}h$ \\
tails & $d \backslash_D t$ & $\lambda x.0^c{\cdot}!h$ \\
is & $(d \backslash t)\slash(d \backslash_D t)$ & $\lambda P.\lambda x.P x$ \\
knows & $(e \backslash t)\slash_M t$ &  $\lambda p.\lambda R. {\Diamond}R p$\\
that  & \multicolumn{2}{l}{$(((e \backslash t){\slash_M}t) \backslash (e \backslash t)) \slash t$} \\
& \multicolumn{2}{l}{$\lambda p.\lambda m.\lambda R.{\sim}(m ({\sim}p) R)$} \\
whether  & \multicolumn{2}{l}{$(((e \backslash t){\slash_M} t) \backslash (e \backslash t)) \slash t$} \\
& \multicolumn{2}{l}{$\lambda p.\lambda m.\lambda R.{\sim}(m ({\sim}p) R)$} \\
& \multicolumn{2}{r}{$+ {\sim}(m p R)$} \\
\end{tabular}
\caption{Partial categorial grammar lexicon. The first column has a word form.
  the second column a categorial type, and third column a semantic translation in
a logical language that extends the Epik term language with lambda.}
\label{lexiconfigure}
\end{figure}

Multimodal categories such as $\backslash_D$ and $\backslash_M$ are
used to control the derivation.  For instance the category of {\em
  heads} is $d \backslash_D t$.  The dummy expletive subject {\em it}
has category $d$.  However the phrase {\em it heads} of category $t$
can not be formed, because $\backslash_D$ is not syntactically active
as a function.  Instead {\em it is heads} is formed with a predicator
{\em is} of category $(d \backslash t)\slash(d \backslash_D t)$.
Similarly {\em knows} has a category with the top-level slash
$\slash\!_M$, and combines to form a sentence as an argument of {\em
  that} or {\em whether}. These complementizers have a category that
looks for the category of {\em know} on the left, after combining with
a complement sentence on the right.

The semantic translations in the third column of Figure
\ref{lexiconfigure} use the Epik term language, incremented with
lambda.  The body of $\lambda x.0^c{\cdot}h$, which is the semantic
lexical entry for {\em heads}, is a term denoting the set of all
worlds where the coin is heads, expressed as the set of all guarded
strings that end with a Boolean valuation where the primitive
proposition $h$ (it's heads) is true.  There is $\lambda x$ at the
front because the grammar formalism enforces a strict correspondence
between syntactic and semantic types. However, the lambda does not
bind anything, because sentences such as {\em it isn't heads} have an
expletive subject.  The body ${\Diamond}R p$ of $\lambda p.\lambda
R. {\Diamond}R p$, which is the semantic lexical entry of {\em knows},
is an Epik term denoting the pre-image of the world-alternative
relation contributed by the subject.  This is not the right semantics
for {\em Amy knows that it's heads}, because it is an existential
modality ${\Diamond_R}p$, rather than an universal modality
${\Box_R}p$.  This is corrected by the complementizer {\em that} or
{\em whether}, which introduces the dual.

Sentences are parsed with a chart parser for categorial grammar.  The
semantics for complex phrases are obtained by syntactic application of
semantic translations, accompanied by beta reduction.  Semantic terms
in the parsing formalism are expressions of untyped lambda calculus.
The grammar is set up so that lambda is eliminated by beta reduction
in the semantic term corresponding to a sentence. In consequence, the
semantic term translating a sentence is a term of the Epik term
language.  Such a term designates a set of possible words (guarded
strings) in the possible worlds model deterimened by an Epik
specification.
By way of example, 
\exi{three}{a} is an English sentence with predicate conjunction and three levels
of clausal embedding.  Using the grammar and parser, the sentence is mapped to the
Epik term \exi{b}, which here takes the form of a Scheme s-expression.

Using the result from Section 3, this term can be mapped in an
implementation of the finite state calculus to a finite state machine that represents
a countably infinite set of possible worlds, represented as guarded strings.  Using
the result from Section 4, it can be mapped to an infinite lazy list of guarded
strings, representing the same set of possible worlds.  Either of these is a concrete
computational representation of the propositional semantic value \semval{Amy knows
  that Bob knows that Amy knows whether it is heads and knows that Bob does not know
  that Amy knows that it is tails}, in the familiar sense of Montague semantics for
natural language.

\begin{example}  \label{three}
Amy knows that Bob knows that Amy knows whether it is heads and
knows that Bob doesnt know that Amy knows that it is tails.
\end{example}

@@

\begin{examples}
\itema Amy knows that it is tails
\itemb {\tt (Not (Diamond amy (Not tails)))}
\end{examples}

\begin{examples}
\itema Bob doesnt know that Amy knows that it is tails
\itemb {\tt (Not (Not (Diamond bob (Not (Not (Diamond amy (Not tails)))))))}
\end{examples}

\begin{examples}
\itema Bob knows that Amy knows whether it is heads.
\itemb {\tt (Not (Diamond bob (Not (Or (Not (Diamond amy (Not heads))) (Not (Diamond amy heads))))))}
\end{examples}

\begin{examples}
\itema Amy knows that Bob knows that Amy knows whether it is heads and knows that Bob doesnt know that Amy knows that it is tails.
\itemb {\tt (And (Not (Diamond amy (Not (Not (Diamond bob (Not (Or (Not (Diamond amy (Not heads))) (Not (Diamond amy heads))))))))) (Not (Diamond amy (Not (Not (Not (Diamond bob (Not (Not (Diamond amy (Not tails))))))))))) }
\end{examples}
